---
title: "CDC Data Training #2"
author: "UCLA Law COVID-19 Behind Bars Data Project"
date: "April 2021"
output: 
  html_document:
    toc: true
    toc_float: true
---

## Introduction

We will cover a number of topics in this training around how to use the UCLA Law COVID-19 Behind Bars Data Project's data for analysis and visualization purposes. You can expect to leave with a better understanding of how to perform comparative analyses amongst carceral facilities along axes such as gender, public/private status, and state. We will also cover how to merge the UCLA data set with external datasets such as the BJS, or Bureau of Justice Statistics, data. 

There is no prior experience using the R programming language required for this training. For ease of use, we recommend opening this R markdown file and following along within the RStudio application.

## Loading Packages

The code below loads in a few R packages that we'll use to manipulate the data in this session. A package bundles together code and documentation, and allows you to access helpful pre-written functions. To learn more about the most widely-used R packages, feel free to visit [this page](https://www.rstudio.com/products/rpackages/).

```{r package setup}
## install packages 
install.packages(c("tidyverse", "devtools")) 
## load packages into session
library(tidyverse)
library(devtools)
```

The UCLA data team wrote a custom R package to help work with our data. You can read more about the package's functionality in the repository homepage, here: https://github.com/uclalawcovid19behindbars/behindbarstools.

```{r}
## install UCLA package
devtools::install_github("uclalawcovid19behindbars/behindbarstools")
## load UCLA package into session
library(behindbarstools) 
## load the UCLA package documentation
help(package=behindbarstools) 
```

## Load Data

- show historical vs latest data

```{r}

```

## Filter Data 

- more than 25 cumulative cases 

## Merge Data 

-- what's useful in BJS, HIFLD? 

## Transform Data

- ??


